Sorting by Reversals

### 2020-07-07 slides.tex
update slides

\begin{definition}
A \b{reversal} $\rho(i,j) \in S_n$ is defined as
\[
    \rho(i,j) = (0\ 1\ \cdots\ i-1\ \r j\ \r{j-1}\ \r \cdots\ \r{i+1}\ \r i\ j+1\ \cdots\ n\ n+1)
\]
for some $j \geq i$.
\end{definition}\pause

@@

\begin{lemma}[\citeauthor*{Christie1998}]
\label{lem:2}
An oriented alternating cycle is a $2$-cycle iff the corresponding reversal acting on both red edges is a $2$-reversal.
\end{lemma}\pause

\begin{proof}
Let $i \sim i'$ and $j \sim j'$.

\begin{center}
\begin{tikzpicture}
	\begin{pgfonlayer}{nodelayer}
		\node [style=node] (0) at (-1.25, 0) {i};
		\node [style=node] (1) at (1.25, 0) {j'};
		\node [style=node] (2) at (3, 0) {i'};
		\node [style=node] (3) at (-3, 0) {j};
		\node (4) at (0, 0) {$\dots$};
		\node (5) at (-4.25, 0) {$\dots$};
		\node (6) at (4.25, 0) {$\dots$};
	\end{pgfonlayer}
	\begin{pgfonlayer}{edgelayer}
		\draw [style=blue edge, bend right] (3) to (1);
		\draw [style=blue edge, bend left] (0) to (2);
		\draw [style=red edge] (3) to (0);
		\draw [style=red edge] (1) to (2);
	\end{pgfonlayer}
\end{tikzpicture}
\end{center}
\end{proof}

@@

\begin{theorem}[lower bound with $2$-cycles, \citeauthor*{Christie1998}]
\label{thm:3}
Let $\pi \in S_n$ and $C$ be the maximum alternating cycle decomposition of $G(\pi)$. Let $c_2(\pi)$ be the minimum number of alternating $2$-cycles in $C$. Then
\begin{align*}
    d(\pi) \geq \frac{2}{3} b(\pi) - \frac{1}{3} c_2(\pi).
\end{align*}
\end{theorem}

@@

Let $u$ be a vertex of $R(C)$.\pause

\begin{definition}
Denote by $C_u$ the alternating cycle decomposition of $G(\pi \circ \rho(u))$ that is obtained from $C$ by removing one 2-cycle or shortening one of its cycles by one red edge.
\end{definition}\pause

@@

\begin{corollary}[\citeauthor*{Christie1998}]
A reversal $\rho(u)$ affects only vertices that are in the same connected component as $u$.
\end{corollary}\pause

\begin{corollary}[\citeauthor*{Christie1998}]
$R(C)$ contains no isolated blue vertices arising from cycles in $C$.
\end{corollary}

@@

\begin{lemma}[\citeauthor*{Christie1998}]
\label{lem:6}
Let $A$ be an oriented connected component of $R(C)$ arising from $k$ different cycles of $G(\pi)$. Then there is an \textit{elimination sequence} for $A$ that contains $k$ $2$-reversals with all the other reversals being $1$-reversals.
\end{lemma}\pause

\begin{proof}
We know that
\begin{itemize}
    \item every vertex in $R(C)$ arising from an alternating cycle of $G(\pi)$ is contained within the same connected component (lemma \ref{lem:3})\pause;
    \item  as $A$ is oriented, we can use lemma \ref{lem:4} to repeatedly find a red vertex $u$ in $A$ and apply $\rho(u)$\pause; and
    \item $\rho(u)$ is a $2$-reversal iff $u$ arises from a $2$-cycle (lemma \ref{lem:2}).
\end{itemize}\pause
Every alternating cycle represented by vertices constituting $A$ eventually reduces to a $2$-cycle.
\end{proof}

@@

@@

\begin{lemma}[\citeauthor*{Christie1998}]
Let $A$ be an unoriented connected component of $R(C)$ arising from $k$ different cycles of $G(\pi)$. Then there is an \textit{elimination sequence} for $A$ that contains one $0$-reversal, $k$ $2$-reversals, with all the other reversals being $1$-reversals.
\end{lemma}\pause

\begin{proof}
Let $\rho(u)$ be a $0$-reversal. Then $A_u$ is oriented (lemma \ref{lem:5}). \par\pause
Now apply lemma \ref{lem:6} to obtain an elimination sequence for $A_u$.
\end{proof}\pause

@@

A selected $2$-cycle is called \b{independent} if it is not part of a ladder. Otherwise it is called a \b{ladder $2$-cycle}.
\end{definition}

@@

\begin{proof}
Let $C$ contain all independent $2$-cycles from $L(M)$ and every alternate cycle of each ladder. \par\pause
Obtain the rest of $C$ by adding any alternating cycle decomposition of the remaining unused edges of $G(\pi)$.
\end{proof}

@@

\begin{proof}[Proof (cont.)]
We know that
\begin{enumerate}
    \item $k+l+m \geq \left\lceil \frac{y}{2} \right\rceil + z$ as $\left\lceil \frac{y}{2} \right\rceil + z$ is the number of selected $2$-cycles by $M$\pause;
    \item $v \leq \left\lceil \frac{z}{2} \right\rceil$ as every unoriented component representing a $2$-cycle represents at least one more alternating cycle (lemma \ref{lem:7})\pause; and
    \item $|M| = y+z \geq c_2(\pi)$.
\end{enumerate}\pause
\begin{align*}
    k+l+m-v &\geq \left\lceil \frac{y}{2} \right\rceil + z - v                                          & (1) \\ \pause
            &\geq \left\lceil \frac{y}{2} \right\rceil + z - \left\lceil \frac{z}{2} \right\rceil       & (2) \\ \pause
            &= \left\lceil \frac{y}{2} \right\rceil + \left\lceil \frac{z}{2} \right\rceil \\ \pause
            &\geq \frac{c_2(\pi)}{2}                                                                    & (3)
\end{align*}
\end{proof}

@@

\printbibliography

>>>>

\begin{definition}
A \b{reversal} $\rho(i,j) \in S_n$ is defined as
\[
    \rho(i,j) = (0\ 1\ \cdots\ i-1\ \r j\ \r{j-1}\ \r \cdots\ \r{i+1}\ \r i\ j+1\ \cdots\ n\ n+1)
\]
for some $i, j \in [1, n]$ with $j \geq i$.
\end{definition}\pause

@@

\begin{lemma}[\citeauthor*{Christie1998}]
\label{lem:2}
An oriented alternating cycle is a $2$-cycle iff the corresponding reversal acting on both red edges is a $2$-reversal.
\end{lemma}\spadding

Let $i \sim i'$ and $j \sim j'$.

\begin{figure}
\begin{center}
\begin{tikzpicture}
	\begin{pgfonlayer}{nodelayer}
		\node [style=node] (0) at (-1.25, 0) {i};
		\node [style=node] (1) at (1.25, 0) {j'};
		\node [style=node] (2) at (3, 0) {i'};
		\node [style=node] (3) at (-3, 0) {j};
		\node (4) at (0, 0) {$\dots$};
		\node (5) at (-4.25, 0) {$\dots$};
		\node (6) at (4.25, 0) {$\dots$};
	\end{pgfonlayer}
	\begin{pgfonlayer}{edgelayer}
		\draw [style=blue edge, bend right] (3) to (1);
		\draw [style=blue edge, bend left] (0) to (2);
		\draw [style=red edge] (3) to (0);
		\draw [style=red edge] (1) to (2);
	\end{pgfonlayer}
\end{tikzpicture}
\end{center}
\caption{Oriented $2$-cycle}
\end{figure}

\begin{figure}
\begin{center}
\begin{tikzpicture}
	\begin{pgfonlayer}{nodelayer}
		\node [style=node] (0) at (-1.25, 0) {i};
		\node [style=node] (1) at (1.25, 0) {i'};
		\node [style=node] (2) at (3, 0) {j'};
		\node [style=node] (3) at (-3, 0) {j};
		\node (4) at (0, 0) {$\dots$};
		\node (5) at (-4.25, 0) {$\dots$};
		\node (6) at (4.25, 0) {$\dots$};
	\end{pgfonlayer}
	\begin{pgfonlayer}{edgelayer}
		\draw [style=blue edge, bend right] (2) to (3);
		\draw [style=blue edge, bend left] (0) to (1);
		\draw [style=red edge] (3) to (0);
		\draw [style=red edge] (1) to (2);
	\end{pgfonlayer}
\end{tikzpicture}
\end{center}
\caption{Unoriented $2$-cycle}
\end{figure}

@@

\begin{theorem}[lower bound with $2$-cycles, \citeauthor*{Christie1998}]
\label{thm:3}
Let $\pi \in S_n$ and $C$ be a maximum alternating cycle decomposition of $G(\pi)$. Let $c_2(\pi)$ be the minimum number of alternating $2$-cycles in any such $C$. Then
\begin{align*}
    d(\pi) \geq \frac{2}{3} b(\pi) - \frac{1}{3} c_2(\pi).
\end{align*}
\end{theorem}

@@

Let $u$ be a vertex of $R(C)$ representing reversal $\rho(u)$.\pause

\begin{definition}
Denote by $C_u$ the alternating cycle decomposition of $G(\pi \circ \rho(u))$ that is obtained from $C$.% by removing one 2-cycle or shortening one of its cycles by one red edge.
\end{definition}\pause

@@

% \begin{corollary}[\citeauthor*{Christie1998}]
% A reversal $\rho(u)$ affects only vertices that are in the same connected component as $u$.
% \end{corollary}\pause

% \begin{corollary}[\citeauthor*{Christie1998}]
% $R(C)$ contains no isolated blue vertices arising from cycles in $C$.
% \end{corollary}

@@

\begin{exampleblock}{Example (elimination sequence)}
Let $\pi = (0\ 1\ 3\ 4\ 2\ 5) \in S_4$\onslide<2->{ and $u_1 = \{1,2\}$}\onslide<3->{, $u_2 = \{2,3\}$.}\spadding

\only<1>{\begin{center}\begin{tikzpicture}
	\begin{pgfonlayer}{nodelayer}
		\node [style=node] (0) at (-3, 0) {0};
		\node [style=node] (1) at (-1.5, 0) {1};
		\node [style=node] (2) at (0, 0) {3};
		\node [style=node] (3) at (1.5, 0) {4};
		\node [style=node] (4) at (3, 0) {2};
		\node [style=node] (5) at (4.5, 0) {5};
	\end{pgfonlayer}
	\begin{pgfonlayer}{edgelayer}
		\draw [style=red edge] (1) to (2);
		\draw [style=red edge] (3) to (4);
		\draw [style=red edge] (4) to (5);
		\draw [style=blue edge, bend right] (1) to (4);
		\draw [style=blue edge, bend left] (2) to (4);
		\draw [style=blue edge, bend left] (3) to (5);
	\end{pgfonlayer}
\end{tikzpicture}\end{center}}\spadding
\only<1>{\begin{center}\begin{tikzpicture}
	\begin{pgfonlayer}{nodelayer}
		\node [style=red node] (0) at (0, 1) {$\{2,3\}$};
		\node [style=red node] (1) at (-2, 1) {$\{1,2\}$};
		\node [style=blue node] (2) at (-1, -1) {$\{4,5\}$};
		\node [style=blue node] (3) at (2, -1) {$\{3,4\}$};
		\node [style=blue node] (4) at (2, 1) {$\{0,1\}$};
	\end{pgfonlayer}
	\begin{pgfonlayer}{edgelayer}
		\draw (0) to (2);
		\draw (1) to (2);
	\end{pgfonlayer}
\end{tikzpicture}\end{center}}
\only<2>{\begin{center}\begin{tikzpicture}
	\begin{pgfonlayer}{nodelayer}
		\node [style=node] (0) at (-3, 0) {0};
		\node [style=node] (1) at (-1.5, 0) {1};
		\node [style=node] (2) at (0, 0) {2};
		\node [style=node] (3) at (1.5, 0) {4};
		\node [style=node] (4) at (3, 0) {3};
		\node [style=node] (5) at (4.5, 0) {5};
	\end{pgfonlayer}
	\begin{pgfonlayer}{edgelayer}
		\draw [style=red edge] (2) to (3);
		\draw [style=red edge] (4) to (5);
		\draw [style=blue edge, bend left] (2) to (4);
		\draw [style=blue edge, bend left] (3) to (5);
	\end{pgfonlayer}
\end{tikzpicture}\end{center}}\spadding
\only<2>{\begin{center}\begin{tikzpicture}
	\begin{pgfonlayer}{nodelayer}
		\node [style=red node] (0) at (0, 1) {$\{2,3\}$};
		\node [style=blue node] (1) at (-2, 1) {$\{1,2\}$};
		\node [style=red node] (2) at (-1, -1) {$\{4,5\}$};
		\node [style=blue node] (3) at (2, -1) {$\{3,4\}$};
		\node [style=blue node] (4) at (2, 1) {$\{0,1\}$};
	\end{pgfonlayer}
	\begin{pgfonlayer}{edgelayer}
		\draw (0) to (2);
	\end{pgfonlayer}
\end{tikzpicture}\end{center}}
\only<3>{\begin{center}\begin{tikzpicture}
	\begin{pgfonlayer}{nodelayer}
		\node [style=node] (0) at (-3, 0) {0};
		\node [style=node] (1) at (-1.5, 0) {1};
		\node [style=node] (2) at (0, 0) {2};
		\node [style=node] (3) at (1.5, 0) {3};
		\node [style=node] (4) at (3, 0) {4};
		\node [style=node] (5) at (4.5, 0) {5};
	\end{pgfonlayer}
\end{tikzpicture}\end{center}}\spadding
\only<3>{\begin{center}\begin{tikzpicture}
	\begin{pgfonlayer}{nodelayer}
		\node [style=blue node] (0) at (0, 1) {$\{2,3\}$};
		\node [style=blue node] (1) at (-2, 1) {$\{1,2\}$};
		\node [style=blue node] (2) at (-1, -1) {$\{4,5\}$};
		\node [style=blue node] (3) at (2, -1) {$\{3,4\}$};
		\node [style=blue node] (4) at (2, 1) {$\{0,1\}$};
	\end{pgfonlayer}
\end{tikzpicture}\end{center}}
\end{exampleblock}

@@

% \begin{frame}

% \begin{lemma}[\citeauthor*{Christie1998}]
% \label{lem:6}
% Let $A$ be an oriented connected component of $R(C)$ arising from $k$ different alternating cycles of $G(\pi)$. Then there is an \textit{elimination sequence} for $A$ that contains $k$ $2$-reversals with all the other reversals being $1$-reversals.
% \end{lemma}\pause

% \begin{proof}
% We know that
% \begin{itemize}
%     \item every vertex in $R(C)$ arising from an alternating cycle of $G(\pi)$ is contained within the same connected component (lemma \ref{lem:3})\pause;
%     \item  as $A$ is oriented, we can use lemma \ref{lem:4} to repeatedly find a red vertex $u$ in $A$ and apply $\rho(u)$\pause; and
%     \item $\rho(u)$ is a $2$-reversal iff $u$ arises from a $2$-cycle (lemma \ref{lem:2}).
% \end{itemize}\pause
% Every alternating cycle represented by vertices constituting $A$ eventually reduces to a $2$-cycle.
% \end{proof}

% \end{frame}

@@

% \begin{lemma}[\citeauthor*{Christie1998}]
% Let $A$ be an unoriented connected component of $R(C)$ arising from $k$ different alternating cycles of $G(\pi)$. Then there is an \textit{elimination sequence} for $A$ that contains one $0$-reversal, $k$ $2$-reversals, with all the other reversals being $1$-reversals.
% \end{lemma}\pause

% \begin{proof}
% Let $\rho(u)$ be a $0$-reversal. Then $A_u$ is oriented (lemma \ref{lem:5}). \par\pause
% Now apply lemma \ref{lem:6} to obtain an elimination sequence for $A_u$.
% \end{proof}\pause

Observations:
\begin{itemize}
    \item Every connected component $A$ arising from k different alternating cycles of $G(\pi)$, eventually reduces to a two cycle.\pause
    \item There exists an elimination sequence of $A$ with $k$ $2$-reversals and remaining $1$-reversals.\pause
    \item An unoriented connected component requires one initial $0$-reversal.
\end{itemize}\spadding\pause

@@

A selected $2$-cycle is called \b{independent} if it is not part of a ladder. Otherwise it is called a \b{ladder $2$-cycle}.
\end{definition}\pause\padding

@@

% \begin{proof}
% Let $C$ contain all independent $2$-cycles from $L(M)$ and every alternate cycle of each ladder. \par\pause
% Obtain the rest of $C$ by adding any alternating cycle decomposition of the remaining unused edges of $G(\pi)$.
% \end{proof}

@@

\begin{proof}[Proof (cont.)]
To show: $k+l+m-v \geq \frac{c_2(\pi)}{2}$. \pause We know that
\begin{enumerate}
    \item $k+l+m \geq \left\lceil \frac{y}{2} \right\rceil + z$ as $\left\lceil \frac{y}{2} \right\rceil + z$ is the number of selected $2$-cycles by $M$\pause;
    \item $v \leq \left\lfloor \frac{z}{2} \right\rfloor$ as every unoriented component representing a $2$-cycle represents at least one more alternating cycle (lemma \ref{lem:7})\pause; and
    \item $|M| = y+z \geq c_2(\pi)$.
\end{enumerate}\pause
\begin{align*}
    k+l+m-v &\geq \left\lceil \frac{y}{2} \right\rceil + z - v                                          & (1) \\ \pause
            &\geq \left\lceil \frac{y}{2} \right\rceil + z - \left\lfloor \frac{z}{2} \right\rfloor     & (2) \\ \pause
            &= \left\lceil \frac{y}{2} \right\rceil + \left\lceil \frac{z}{2} \right\rceil \\ \pause
            &\geq \frac{c_2(\pi)}{2}                                                                    & (3)
\end{align*}
\end{proof}

Run time: $O(n^4)$, can be improved to $O(n^2)$.\spadding

\begin{block}{Summary}
\begin{itemize}
    \item the number of alternating cycles in a breakpoint graph $G(\pi)$ are related to $d(\pi)$
    \item a sorting sequence of reversals can be constructed from an alternating cycle decomposition of $G(\pi)$
\end{itemize}
\end{block}

\begin{block}{Outlook}
\begin{itemize}
    \item there exists a $1.375$-approximation (\citeauthor*{Berman2001})
    \item MIN-SBR for signed permutations is in $P$ (\citeauthor*{Hannenhalli1995})
    \item MIN-SBR for signed permutations is in $P$ (\citeauthor*{Hannenhalli1995})
\end{itemize}
\end{block}

@@

\nocite{*}
\printbibliography
